---
title: "Lice at the limits: impact of regulatory limits on lice counts in salmon farms"
author:
  - name: Richard J Telford
    orcid: 0000-0001-9826-3076
    email: richard.telford@uib.no
    affiliations:
      - name: Department of Biological Sciences, University of Bergen 
        city: Bergen
        address: Postbox 7803
        country: Norway
        postal-code: N-5020
  - name: et qui?
format: 
  pdf:
    papersize: a4
  html: 
    self-contained: true
editor: visual
abstract: >
  Analysis of sea lice counts show unexpected features near to regulatory limits 
  on the number of lice per fish. 
  These features are consistent with aquaculture facilities taking advantage of 
  loopholes in the regulations to penalties associated with exceeding those limits.
bibliography:
  - pakret.bib
  - lice.bib
#csl: aquaculture-international.csl
execute:
  echo: false
  error: true
---

```{r}
#| label: setup
#| message: false

# packages used in the document
library(targets)
library(pakret)
library(gt)
library(ggplot2) # just to set themes


# set default ggplot theme
theme_set(theme_bw())
```

## Introduction

Regulations impose a burden on the regulated. 
There is a cost to compliance that companies will seek to minimise by developing efficient procedures, by circumventing the regulations, or by breaching them if the penalties are low or enforcement is weak.
If circumvention is widespread, regulations may fail to reach their objectives even though all parties are apparently fully complying.

Polluting industries are often regulated by imposing limits emissions. [acid rain, CO2, nuclear discharge, sewerage works, etc etc].
A special case of a polluting industry is the aquaculture industry. 
In cool temperate and sub-polar coastal waters, salmon farms are a major economic concern with worldwide revenues of $xxxx #. 
Most farmed salmon are raised in open net pens which release pharmaceutical products and organic waste into the water #. 
These emissions are regulated #.

In addition, the farmed salmon host _Lepeophtheirus salmonis_, the salmon louse, an ectoparasitic copepod that infests salmonoid fish causing welfare problems# and economic losses# in the aquaculture industry. 
The farms release lice eggs into the environment where they can infest wild salmonoid fish#. 
Salmon lice released by the aquaculture industry is thought to be one of the drivers of decreasing wild salmon populations# alongside climate change xxx #.

Consequently, countries with salmon farming have established regulatory frameworks to control lice infestations [@Vormedal2023]. 
These regulations focus on the mean number of adult lice per fish [@tbl-regulations] which are measured regularly and reported to the regulators.
In most jurisdictions, these louse counts are publicly available.

```{r}
#| label: tbl-regulations
#| tbl-cap: "Regulatory limits on the number of adult lice per fish."
tibble::tribble(
  ~Country, ~Period, ~"Limit per fish", 
  "British Columbia, Canada", "????", "3 motile lice", 
  "Ireland", "2008-", "0.5 egg-bearing lice during the smolt out-migration, 2 otherwise",
  "Norway", "2009-2012", "1 adult female lice, or 5 motile lice", 
  "Norway", "2013-2016", "0.5 adult female lice", 
  "Norway", "2017-", "0.2 adult female lice during smolt out-migration, 0.5 otherwise", 
  "Scotland", "2021-", "2 adult female lice triggers increases monitoring, 6 requires treatment") |> 
  gt() |> 
  cols_width(starts_with("Limit") ~ px(260))
```

Jeong et al. [-@Jeong2023] reported that in Norway, but not in British Columbia, the distribution of louse counts has an unexpected "cliff" at the regulatory limit.
Jeong et al. [-@Jeong2023] ascribe this "cliff" to a desire by aquaculture managers to report values below the regulatory limits, but do not identify the mechanisms by which they could do this. 
Sævik and Sandvik [-@Saevik2023] dispute this interpretation, arguing that the "cliff" is due to efficient lice treatment in Norway.

In this paper, we aim to extend the analysis of Jeong et al. [-@Jeong2023] to other countries with farmed and wild salmonoids.
We test if there is a "cliff" or discontinuity in the louse counts distribution at the regulatory limit and try to explain it by natural processes or due to the regulations being circumvented.


## Methods

```{r}
#| label: period

tar_load(periods)

```


We examine the distribution of the mean louse counts reported by the aquaculture facilities to the national regulators using publicly available data.
Lice count data were downloaded from the regulators in British Columbia, (Canada) (`r periods$britishColumbia`) [#], Ireland (`r periods$ireland`) [#], Norway (`r periods$norway`) [#barentswatch], and Scotland (`r periods$scotland`) [#].

We test for discontinuities at the regulatory limits using the `r pkrt("rddensity")` package.
All analyses were done in `r pkrt("R")`. 
Code to reproduce the analyses is available from https://github.com/richardjtelford/lice-at-the-limit




# Results




```{r}
#| label: fig-hist
#| fig-cap: Histograms of louse counts from selected periods in Scotland and Norway. Histograms are truncated at 3 lice per fish. Vertical dashed lines show the applicable regulatory limits. Histogram bins are 0.01 lice wide. Counts with zero lice are omitted.
#| fig-height: 5
#| warning: false

tar_read(lice_hist)
```



The Scottish data and the Norwegian data from 2012 shows a heavy-tailed, monotonic decline in louse counts [@fig-hist].
The spikes in the data either reflect rounding or many counts being based on 20 fish.
The spikes do not affect the rest of the analysis.
Norwegian louse counts post-2017 show a different pattern, with a sharp drop to near zero counts at the regulatory limit.
The data from outside the smolt migration period show a small rise in counts just below the regulatory limit.
Histograms from 2013-2016 are transitional between the patterns shown in 2012 and post-2017.

```{r}
#| label: load-stats-text
tar_load(stats_text)
```



Density discontinuity analysis [@Cattaneo2018] finds no evidence of a discontinuity at the regulatory limit in British Columbia (`r stats_text$britishColumbia`) ([@fig-rdd]A) or in Ireland  (`r stats_text$ireland`) during the smolt migration season. 
The Scottish lice data show a significant discontinuity at the regulatory limit (`r stats_text$scotland`) ([@fig-rdd]B). 
There is no evidence of a discontinuity at the regulatory limit in Norway in 2012 (`r stats_text$n_2012`) ([@fig-rdd]C), but a discontinuity develops in subsequent years [@fig-t-stat] and is large in, for example, 2020 (`r stats_text$n_2020`) ([@fig-rdd]D).

```{r}
#| label: fig-rdd
#| fig-cap: >
#|   Density discontinuity plots for A) British Columbia (2011--2024), B) Scotland (2021-2023), C) Norway 2012, and D) Norway 2020 outside the migration period.  
#|   Green bars show a histogram of louse counts, black and red lines show the estimated density below and above the regulatory limit respectively, with the uncertainty as a shaded area.  

tar_read(rdds_plots)


```

```{r}
#| label: fig-t-stat
#| fig-cap: t-statistics for the density discontinuities at the regulatory limits in Norway for different years. From 2017, the smolt migration season has a lower limit and is shown separately. [#years with positive t-values in have badly fitting models - might be able to fix these#]

tar_read(norway_t_plot)
```



## Discussion

### Expected distribution of lice counts

The louse count distribution is the result of combining many different distributions for louse counts at different temperatures (fewer lice at colder temperatures), and different fish ages (more lice on older fish), and time since treatment to remove lice.
The exact shape this combined distribution should take is not known, but the heavy-tailed monotonic decline shown by data from British Columbia, Scotland, and Norway in 2012 seems like a plausible distribution.

### Explaining the unexpected patterns in the Norwegian data

The patterns in the Norwegian data are unexpected. 
There are at least four explanations for the patterns.

First, as argued by Sævik and Sandvik [-@Saevik2023] in response to Jeong et al. [-@Jeong2023], the sharp drop in lice numbers at the regulatory limit could be due to frequent and effective de-lousing treatments employed by the Norwegian aquaculture industry.
Such treatments will undoubtedly steepen the curve of the lice count distribution, especially close to and above the regulatory limit, but cannot explain the abrupt drop at the limit because counting uncertainties would result in a smoother decline. 
This mechanism also does not explain the absence of a discontinuity in the Irish lice counts, where the regulatory limits are similar to Norway.
The effect of counting uncertainty can be shown by simulating lice counts when the true counts are all strictly below the limit (@fig-count-uncertainty).
The simulation shows that counting uncertainty will mean that, even in this perfect case, there will not be an abrupt cliff in lice counts and many lice counts will be above the regulatory limit. 
If the number of lice per fish was simulated with a negative binomial, to allow for over-dispersion, or the fewer fish were inspected, the slope would be even shallower.

[I think the simulation in Saevik2023 cleans immediately the lice count reaches 0.5 lice/fish and then reports after cleaning.]

```{r}
#| label: fig-count-uncertainty
#| fig-cap: Lice counts simulated with a Poisson distribution from 10000 surveys with 100 fish examined in each survey. The true lice counts are uniformly distributed between 0 and the regulatory limit of 0.5 lice per fish (denoted with the dashed line). Note that the density discontinuity test does not detect a significant discontinuity in the simulated data. 

tar_read(count_uncertainty)
```


```{r}
#| label: optimal-stopping
tar_load(optimal_stopping)
```


Second, when a count is above the regulatory limit, the facility could choose to count lice on more fish in the hope that the mean count will drop below the limit.
This could be portrayed by managers as an attempt to get more accurate data.
However, if the counts would not be increased when the count was slightly below the regulatory limit, the process with bias the lice count.
This would lead to a increase in the number of counts just below the regulatory limit, as observed. 
Norwegian regulations [@sealicelaw2012] require that counts are based on _at least_ twenty fish per pen (ten outside of the smolt migration period). 
Counting extra fish if the louse count is above the regulatory limit is not prohibited.
A correspondent within the Norwegian aquaculture industry confirmed that some companies use this procedure, but that it is rarely effective.
It is akin to optional stopping in hypothesis testing, where the sample size is increased if statistical tests are not significant. 
Optional stopping is a form of p-hacking recognised as a questionable research practice [@Stefan2023] and causes an excess of p-values just below the significance threshold, and a deficit just above.
The severity of optional stopping for gaming limits depends on the number of times the current count is checked and the number of fish examined between checks [@Stefan2023]. 

To show the effect of optimal stopping, we run a simulation of `r optimal_stopping$nsim` lice counts. In each count, the lice have a Poisson distribution with a mean of `r optimal_stopping$true_mean` per fish. If the original count on `r optimal_stopping$noriginal` fish exceeded the regulatory limit of `r optimal_stopping$threshold` lice per fish, an extra `r optimal_stopping$nextra` fish are counted and the mean louse count recalculated.
In a simulation `r optimal_stopping$thresh_original`% of the original counts exceed the limit, whereas `r optimal_stopping$thresh_final`% of the counts after optimal stopping exceed the limit. 

Third, facilities with a count above the regulatory limit could immediately treat the fish to remove lice, and then recount the lice, and report the lower post-cleaning count.
This would lead to a drop in louse counts at the legal limit. 
The abrupt drop in louse counts suggests that optional stopping or reporting post-treatment counts are widespread in the Norwegian aquaculture industry to the extent that they are standard practice.
Norwegian regulations require a louse count to be reported each week (every two weeks when the temperature is below 4˚C) [@sealicelaw2012].
Nothing in the regulations prohibits recounting lice after treatment.

Finally, the patterns could be due to data falsification or fabrication. 
Two such cases have reached Norway's supreme court.
In 2015, one company with four facilities in northern Norway and its manager was convicted for fabricating and falsifying louse counts over a six month period in 2012--2013#. 
Although the infestation reached 8.33 adult female lice per fish, the company avoided having to treat the lice by misreporting the louse counts.
In a previous case#, a manager of a company was convicted of reporting that there were no lice when counts had not been carried out.
The prevalence of fabrication and falsification is probably low as the penalties for misreporting counts are high and the aquaculture industry not centralised. 
Thus fabrication and falsification probably explain little of the observed patterns in the Norwegian lice count distribution.

### Implications

Suggests that facilities attempt to comply with the regulations by treating lice when the limit is exceeded rather than treating in anticipation to prevent the limit being exceeded.

[use lice model to investigate difference in apparent infection pressure and true infection pressure with treat-then-report rather than report-then-treat]

Counts of salmon lice are currently the only metric used by the Norwegian Traffic Light System# to balance the growth of the aquaculture industry with the need to protect wild salmon populations.

If the count data do not reflect the true number of lice, estimates of lice infection pressure # will also be biased.

[expand]

Are facilites that count then treat if above limit in compliance with the law?

Norwegian salmon louse counts from #barentswatch are widely used.  Who uses barrentswatch data... check citations. classify into groups

lice infestion pressure, lice-induced mortality of wild salmon

Stige2022 drive their models lice-induced mortality of wild salmon with louse count data. 

### Recommendations

The British Columbia lice counts does not show a discontinuity at the regulatory limit. 
Aquaculture facilities in the province are regularly audited by the regulator #. 
Godwin et al [-@Godwin2020] showed that lice counts are almost twenty percent higher in audits than in routine counts.

- fix count sizes to prevent optional stopping
- require that all counts are reported
- spot checks to validate counts

## Conclusions

Counts of salmon lice from Norwegian aquaculture facilities do not fully reflect the true number of lice as two procedures are used to artificially deflate the counts when they are above the regulatory limit

## Acknowledgements

So long, and thanks for all the fish.

## Data availability

All data are publicly available from the regulars in Canada, Ireland, Norway, and Scotland. 
Code to reproduce this manuscript is available from  https://github.com/richardjtelford/lice-at-the-limit


## References

https://www.domstol.no/globalassets/upload/hret/avgjorelser/2016/avgjorelser-desember-2016/sak-2016-1391.pdf

https://www.domstol.no/globalassets/upload/hret/avgjorelser/2015/saknr-2014-1689.pdf


A. Elmoslemany, S.K. Whyte, C.W. Revie, K.L. Hammell
Sea lice monitoring on Atlantic salmon farms in New Brunswick, Canada: comparing audit and farm staff counts
J. Fish Dis., 36 (2013), pp. 241-247

::: {#refs}
:::



Skilbrei OT, Heino M, Svåsand T. (2015) Using simulated escape events to assess the annual numbers and
destinies of escaped farmed Atlantic salmon of different life stages, from farms sites in Norway. ICES Journal of
Marine Science, 72 : 670–685.
https://doi.org/10.1093/icesjms/fsu133


Fate and effects of fish farm organic waste in marine systems: Advances in understanding using biochemical approaches with implications for environmental management
Deanna M. Elvines, Catriona K. MacLeod, Donald. J. Ross, Grant A. Hopkins, Camille A. White
First published: 05 April 2023
https://doi.org/10.1111/raq.12821

https://doi.org/10.3354/aei00443 Stige2022
