---
title: "Supplementary material for lice at the limits: circumvention of regulatory limits on salmon lice counts in aquaculture facilities"
author:
  - name: Richard J Telford
    orcid: 0000-0001-9826-3076
    email: richard.telford@uib.no
    affiliations:
      - name: Department of Biological Sciences, University of Bergen 
        city: Bergen
        address: Postbox 7803
        country: Norway
        postal-code: N-5020
format: 
  aft-pdf:
    papersize: a4
    keep-tex: true  
  html: 
    self-contained: false
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Figure S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
editor: visual
execute:
  echo: false
  error: true
---


Histograms showing the distribution of the mean number of lice per fish in British Columbia (@suppfig-bc), Iceland (@suppfig-iceland), Ireland (@suppfig-ireland), and Norway (@suppfig-norway1, @suppfig-norway2) when the lice count is greater than zero.
All histograms are truncated to better show the distribution of lice counts near the regulatory limit.
The vertical dashed line shows the regulatory limit on the number of lice per fish. 
In Ireland and in Norway from 2017, the limit is reduced in the smolt migration season.
There is enough data from Norway to show the distribution of lice counts for each year.

```{r}
#| label: setup
#| message: false

# packages used in the document
library(targets)
library(tidyverse)


# set default ggplot theme
theme_set(theme_bw())

# load data
tar_load(all_data)
```

```{r}
#| label: hist-function

plot_hist <- function(data, x, xlab = "Mean number of adult female lice per fish", xlim = c(0, 3), binwidth = 0.05, vline = NULL, facet = "none") {
  p <- data |> 
    filter(lice > 0) |> 
    ggplot(aes(x = lice)) +
    geom_histogram(boundary = -.Machine$double.eps, binwidth = binwidth) +
    coord_cartesian(xlim = xlim) +
    labs(x = xlab, y = "Number of surveys")
  
  if (!is.null(vline)) {
    p <- p + 
      geom_vline(data = vline, aes(xintercept = limit), linetype = "dashed", colour = "#99151c")
  }
  
  if (facet == "year") {
    p <- p + facet_wrap(vars(year), ncol = 2)
  } else if (facet == "seasonyear") {
    p <- p + facet_grid(rows = vars(year), cols = vars(as.character(limit)), 
                        labeller = labeller(limit = c("0.2" = "Smolt migration", "0.5" = "Ex smolt migration")))
  } else if (facet == "season") {
    p <- p + 
      facet_wrap(vars(limit), 
                 labeller = labeller(limit = c("0.5" = "Smolt migration", "2" = "Ex smolt migration")))
  }
  
  p
}
```


:::{#suppfig-bc}
```{r}

all_data$britishColumbia |> 
  plot_hist(vline = tibble(limit = 3), xlim = c(0, 5), 
            xlab = "Mean number of motile lice per fish")

```

British Columbia
:::

:::{#suppfig-iceland}
```{r}

all_data$iceland |> plot_hist(vline = tibble(limit = 0.5))

```

Iceland
:::

:::{#suppfig-ireland}

```{r}
all_data$ireland |> 
  plot_hist(vline =  tibble(limit = c(0.5, 2)),
            xlab = "Mean number of egg-bearing lice per fish",
            facet = "season")
```
Ireland
:::

:::{#suppfig-norway1}

```{r}
#| fig-height: 8 


all_data$norway |> 
  filter(year < 2017) |> 
  plot_hist(vline = tibble(limit = 0.5), facet = "year", xlim = c(0, 2))
```
Norway 2012 -- 2016
:::


:::{#suppfig-norway2}

```{r}
#| fig-height: 12
#| fig-width: 6

all_data$norway |> 
  filter(!limit == "Ukjent") |> 
  filter(year >= 2017) |> 
  plot_hist(vline = tibble(limit = c(0.2, 0.5)), facet = "seasonyear", xlim = c(0, 1))
```
Norway since 2017
:::
